---
title: Data Science - Reproducible Research - Peer Assessment 2 - Analysis of Weather Event in US and Harmful and Economic impact using NOAA Storm Database 
author: "by jmvilaverde"
date: "Monday, June 15, 2015"
output:
  html_document:
    keep_md: yes
    pandoc_args:
    - +RTS
    - -K64m
    - -RTS
---




***


##Synopsis: 

Based on data extracted from NOAA Storm Database for event registered in US from year 1996 to present, and analyzed with the procedure described in next section _Data Processing_, are obtained the subsequent conclusions:

Human harmful:
* *TOP 3* most harmful events are *TORNADO (33.9%), HEAT (14.3%) and FLOOD (11%)* that causes *59.2 %* of total harmful of all events. 
* *TOP 10* causes *86 %* of total harmful, it is a total of *56179 fatalities or people injured*.

Economic damages:
* *TOP 3* event types that causes most economic damages are *FLOOD (37.4%), HURRICANTE/TYPHOON (18%) and STORM SURGE (10.9%)* that causes *66.3 %* of total Economic damage of all events. 
* *TOP 10* causes *91.8 %* of total economic damages, over *397 Billion $*.

_Can view detail in section Results._

***

##Data Processing

####1.First step, adquire the date from [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2] and put it into a data frame container:

```{r dataAdquisition, echo=TRUE, cache=TRUE}
#Set data path, using setInternet2 to avoid problems with https download
setInternet2(use = TRUE)
dataPath <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- "StormData.csv.bz2"

#Download the file
if (!file.exists(fileName)) download.file(url = dataPath, destfile = fileName)

#Extract bz2 file to a dataframe
dataInitial <- read.csv(bzfile(fileName))
```

As additional information, on http://www.ncdc.noaa.gov/stormevents/details.jsp?type=eventtype
inform about information recolection, and only have a complete collection from 1996 to present:
_3. All Event Types (48 from Directive 10-1605): From 1996 to present, 48 event types are recorded as defined in NWS Directive 10-1605._

Because only have a complete collection from 1996 to present is taken as criteria to filter the data to keep only information from 1996 to present.

####2.Get only relevant fields:

```{r echo=TRUE}
#Identify fields
names(dataInitial)
```

For our process are needed: BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.

Are relevant for filter Date (BGN_DATE) and States (STATE).

####3.Analyze BGN_DATE and STATE, in order to know what class have the data and how to process it: 

```{r echo=TRUE}
#Identify structure
str(dataInitial$BGN_DATE)
head(dataInitial$BGN_DATE)
str(dataInitial$STATE)
```

BGN_DATE and STATE are factors.

####4.We need to transform BGN_DATE from factor into a Date class in order to filter by year.

```{r filterDate, echo=TRUE}

#Use library lubridate to manage Date
library(lubridate)

#Transform BGN_DATE into a class  Date
dataInitial$date <- as.Date(dataInitial$BGN_DATE, format = "%m/%d/%Y")
class(dataInitial$date)

#Data filtered by year >= 1996
dataYearProcessed <- dataInitial[year(dataInitial$date)>=1996,]

#Percent of data removed filtering by Year
percentDataRemYear <- round((1-(nrow(dataYearProcessed)/nrow(dataInitial))) * 100, 2)
```

This filter has removed `r percentDataRemYear`% of data.

####5.Filter to get only data from US States:

```{r filterStates, echo=TRUE}
#Data filtered by States
dataStatesProcessed <- dataYearProcessed[dataYearProcessed$STATE %in% state.abb,]

#Percent of data removed filtering by State
percentDataRemStates <- round((1- (nrow(dataStatesProcessed)/nrow(dataYearProcessed)))*100, 2)
```

This filter has removed `r percentDataRemStates`% of data from precedent filter.


####6.Select necesary fields: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP.

```{r processFilterData, echo=TRUE}
library(dplyr)

#Select only necesary fields: Type of event, Number of fatalities, Number of injuries, Property damages, Property damages exponent,
#Crop damages, and Crop damages exponent.
dataStatesProcessed %>% 
        select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) -> dataPreprocessed
```

####7.Evaluate event types and clean. Do this by unifying and removing event types:

Use [http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf] as reference to clean types.

* Unify TSTM WIND with THUNDERSTORM WIND.
* Unify EXCESSIVE HEAT with HEAT.
* Remove Summary, that is not an Event Type.

```{r echo=TRUE}
#Evaluate how many levels have the dataset
unique(dataPreprocessed$EVTYPE)

##Running the report, are detected some Events that are repeated in the top 10 table.
#TSTM WIND it's the same as THUNDERSTORM WIND. We make the sustitution on data
dataPreprocessed[grepl(pattern = "TSTM WIND", x=dataPreprocessed$EVTYPE),]$EVTYPE <- "THUNDERSTORM WIND"

#EXCESSIVE HEAT it's the same as HEAT. We make the sustitution on data
dataPreprocessed[grepl(pattern = "EXCESSIVE HEAT", x=dataPreprocessed$EVTYPE),]$EVTYPE <- "HEAT"

#Remove Summary from event type
dataPreprocessed <- dataPreprocessed[!grepl(pattern = "Summary", x=dataPreprocessed$EVTYPE),]
```

####8.Analyze content of PROPDMGEXP and CROPDMGEXP:

```{r echo=TRUE}
summary(dataPreprocessed$PROPDMGEXP)
summary(dataPreprocessed$CROPDMGEXP)
```

Use as EXP values B, M and K.

####9.Create new columns: 

* HealthDamage = FATALITIES + INJURIES
* Transform PROPDMGEXP and CROPDMGEXP into a number: B -> 10^9, M -> 10^6, K -> 10^3
* PropertyDamage = PROPDMG * newPROPDMGEXP
* CropDamage = CROPDMG * newCROPDMGEXP

```{r echo=TRUE}
#Create Health harm total
dataPreprocessed %>%
         mutate(HealthDamage = FATALITIES + INJURIES) -> dataPreprocessed

#Create Property damage total
#Type of exponents -> Billion, Million, K thousand
dataPreprocessed %>%
         mutate(multiPROPDM = ifelse(PROPDMGEXP == 'B', 10^9, 
                                     ifelse(PROPDMGEXP == 'M', 10^6, 
                                            ifelse(PROPDMGEXP == 'K', 10^3, 1)))) %>%
         mutate(PropDamage = as.numeric(multiPROPDM) * as.numeric(PROPDMG)) -> dataPreprocessed

#Create Crop damage total
#Type of exponents -> Billion, Million, K thousand
dataPreprocessed %>%
         mutate(multiCROPDM = ifelse(CROPDMGEXP == 'B', 10^9, 
                                     ifelse(CROPDMGEXP == 'M', 10^6, 
                                            ifelse(CROPDMGEXP == 'K', 10^3, 1)))) %>%
         mutate(CropDamage = as.numeric(multiCROPDM) * as.numeric(CROPDMG)) -> dataPreprocessed

#Create Damage total
dataPreprocessed %>%
         mutate(EconomicDamage = PropDamage + CropDamage) -> dataPreprocessed

summary(dataPreprocessed)
```


####10.Create TOP10 data sets Harmful and Economic damages:

```{r getFatalitiesAndInjuries, echo=TRUE}
##TOP per Event Type
#Set top to 10
top <- 10

##TOP of Combined Health Harmful per Event Type
#Group by EVTYPE and sum all HealthDamage, rename columns and order desc by HealthDamage
dataAgregateCombined <- with(dataPreprocessed, aggregate(HealthDamage, list(EVTYPE), sum))
dataAgregateCombined <- rename(dataAgregateCombined, EventType = Group.1, HealthDamage = x)
dataAgrCombinedTop <- arrange(dataAgregateCombined, desc(HealthDamage))[1:top,]

#Change the factor order
dataAgrCombinedTop$EventType <- factor(dataAgrCombinedTop$EventType, levels = dataAgrCombinedTop$EventType[order(dataAgrCombinedTop$HealthDamage)])

#Calculate total Health damage and total Health damage per TOP10 and TOP3
totalHealthDamage <- sum(dataAgregateCombined$HealthDamage)
total.HealthDamage.Top10 <- sum(dataAgrCombinedTop$HealthDamage)
total.HealthDamage.Top3 <- sum(dataAgrCombinedTop$HealthDamage[1:3])

#Calculate percent of Health damage per Event Type
dataAgrCombinedTop %>% mutate(percentHealthDamage = round(HealthDamage / totalHealthDamage * 100,1)) -> dataAgrCombinedTop


##TOP of Fatalities per Event Type
#Group by EVTYPE and sum all FATALITIES, rename columns and order desc by FATALITIES
dataAgregateFatalities <- with(dataPreprocessed, aggregate(FATALITIES, list(EVTYPE), sum))
dataAgregateFatalities <- rename(dataAgregateFatalities, EventType = Group.1, FATALITIES = x)
dAFatalitiesTop <-arrange(dataAgregateFatalities,desc(FATALITIES))[1:top,]

#Change the factor order
dAFatalitiesTop$EventType <- factor(dAFatalitiesTop$EventType, levels = dAFatalitiesTop$EventType[order(dAFatalitiesTop$FATALITIES)])

#Calculate total FATALITIES 
totalFatalities <- sum(dataAgregateFatalities$FATALITIES)

#Calculate percent of FATALITIES  per Event Type
dAFatalitiesTop %>% mutate(percentFatalities = round(FATALITIES / totalFatalities * 100,1)) -> dAFatalitiesTop


##TOP of Injuries per Event Type
#Group by EVTYPE and sum all INJURIES, rename columns and order desc by INJURIES
dataAgregateINJURIES <- with(dataPreprocessed, aggregate(INJURIES, list(EVTYPE), sum))
dataAgregateINJURIES <- rename(dataAgregateINJURIES, EventType = Group.1, INJURIES = x)
dAInjuriesTop <- arrange(dataAgregateINJURIES, desc(INJURIES))[1:top,]

#Calculate total INJURIES 
totalInjuries <- sum(dataAgregateINJURIES$INJURIES)

#Calculate percent of INJURIES  per Event Type
dAInjuriesTop %>% mutate(percentInjuries = round(INJURIES / totalInjuries * 100,1)) -> dAInjuriesTop

#Change the factor order
dAInjuriesTop$EventType <- factor(dAInjuriesTop$EventType, levels = dAInjuriesTop$EventType[order(dAInjuriesTop$INJURIES)])
```

```{r getDamages, echo=TRUE}
##TOP per Event Type
#Set top to 10
top <- 10

##TOP of Combined Economic Damage per Event Type
#Group by EVTYPE and sum all EconomicDamage, rename columns and order desc by EconomicDamage
dataAgregate.EconomicDamage <- with(dataPreprocessed, aggregate(EconomicDamage, list(EVTYPE), sum))
dataAgregate.EconomicDamage <- rename(dataAgregate.EconomicDamage, EventType = Group.1, EconomicDamage = x)
dataAgregate.EconomicDamage.Top <- arrange(dataAgregate.EconomicDamage, desc(EconomicDamage))[1:top,]

#Change the factor order
dataAgregate.EconomicDamage.Top$EventType <- factor(dataAgregate.EconomicDamage.Top$EventType, levels = dataAgregate.EconomicDamage.Top$EventType[order(dataAgregate.EconomicDamage.Top$EconomicDamage)])

#Calculate total Health damage and total Health damage per TOP10 and TOP3
total.EconomicDamage <- sum(dataAgregate.EconomicDamage$EconomicDamage)
total.EconomicDamage.Top10 <- sum(dataAgregate.EconomicDamage$EconomicDamage.Top)
total.EconomicDamage.Top3 <- sum(dataAgregate.EconomicDamage$EconomicDamage.Top[1:3])
total.EconomicDamage.Top10
total.EconomicDamage.Top3
#Calculate percent of Health damage per Event Type
dataAgregate.EconomicDamage.Top %>% 
        mutate(percent.EconomicDamage = round(EconomicDamage / total.EconomicDamage * 100,1)) -> dataAgregate.EconomicDamage.Top


##TOP of Prop Damages per Event Type
#Group by EVTYPE and sum all PropDamage, rename columns and order desc by PropDamage
dataAgregate.PropDamage <- with(dataPreprocessed, aggregate(PropDamage, list(EVTYPE), sum))
dataAgregate.PropDamage <- rename(dataAgregate.PropDamage, EventType = Group.1, PropDamage = x)
dataAgregate.PropDamage.Top <-arrange(dataAgregate.PropDamage,desc(PropDamage))[1:top,]

#Change the factor order
dataAgregate.PropDamage.Top$EventType <- factor(dataAgregate.PropDamage.Top$EventType, levels = dataAgregate.PropDamage.Top$EventType[order(dataAgregate.PropDamage.Top$PropDamage)])

#Calculate total FATALITIES 
total.PropDamage <- sum(dataAgregate.PropDamage$PropDamage)

#Calculate percent of FATALITIES  per Event Type
dataAgregate.PropDamage.Top %>% 
        mutate(percent.PropDamage = round(PropDamage / total.PropDamage * 100,1)) -> dataAgregate.PropDamage.Top


##TOP of Crop Damages per Event Type
#Group by EVTYPE and sum all CropDamage, rename columns and order desc by CropDamage
dataAgregate.CropDamage <- with(dataPreprocessed, aggregate(CropDamage, list(EVTYPE), sum))
dataAgregate.CropDamage <- rename(dataAgregate.CropDamage, EventType = Group.1, CropDamage = x)
dataAgregate.CropDamage.Top <- arrange(dataAgregate.CropDamage, desc(CropDamage))[1:top,]

#Calculate total INJURIES 
total.CropDamage <- sum(dataAgregate.CropDamage$CropDamage)

#Calculate percent of INJURIES  per Event Type
dataAgregate.CropDamage.Top %>% 
        mutate(percent.CropDamage = round(CropDamage / total.CropDamage * 100,1)) -> dataAgregate.CropDamage.Top

#Change the factor order
dataAgregate.CropDamage.Top$EventType <- factor(dataAgregate.CropDamage.Top$EventType, levels = dataAgregate.CropDamage.Top$EventType[order(dataAgregate.CropDamage.Top$CropDamage)])


```

####11. Code to generate plots that shows Top 10 Event type per Fatalities, Injuries and combined of both:

```{r plotHarmful, echo=TRUE, eval=FALSE, fig.height=10, fig.width=11}
library(ggplot2)

#Var to define margin to put labels over bars on plot
barMargin <- 1.2

##Plot combined Harmful 

#Change factor order to plotting
dataAgrCombinedTop$EventType <- factor(dataAgrCombinedTop$EventType, 
                                       levels = dataAgrCombinedTop$EventType[order(dataAgrCombinedTop$HealthDamage)])

# getPlot <- function(dataPlot, xfield, yfield, xlabel, ylabel, title){
#         g <- ggplot(data=dataPlot, aes(x=xfield, y=yfield , fill=xfield))
#         g <- g + geom_bar(stat="identity", show_guide = FALSE, colour="black")
#         g <- g + ylim(0, max(yfield)*1.2)
#         g <- g + coord_flip() + geom_text(aes(label=yfield), hjust=-.25, vjust=0.5, size=4)
#         g <- g + xlab(xlabel) + ylab(ylabel)
#         g <- g + ggtitle(title)
#         g
#         #return ggplot.resultant
# }
# 
# ggplotCombined <- getPlot(dataPlot = dataAgrCombinedTop, xfield = dataAgrCombinedTop$EventType, 
#                           yfield = dataAgrCombinedTop$HealthDamage, xlabel = "Event type",
#                           ylabel = "Combined: Fatalities + Injuries", 
#                           title = "Top 10 Event type per Total Combined: Fatalities + Injuries")

##Common parameters
titleSize <- 10
gb <- geom_bar(stat="identity", show_guide = FALSE, colour="black")
t <- theme(plot.title = element_text(size=titleSize))
#gcommon <- gb + t     



g <- ggplot(data=dataAgrCombinedTop, aes(x=EventType, y=HealthDamage, fill=EventType))
g <- g + gb
g <- g + ylim(0, dataAgrCombinedTop$HealthDamage[1]*barMargin)
g <- g + coord_flip() + geom_text(aes(label=HealthDamage), hjust=-.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab("Number of Fatalities + Injuries") 
g <- g + ggtitle("Total Combined: Fatalities + Injuries per Event Type")
ggplotCombined <- g + t

        g <- ggplot(data=dataAgrCombinedTop, aes(x=EventType, y=percentHealthDamage , fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dataAgrCombinedTop$percentHealthDamage[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percentHealthDamage), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% Combined: Fatalities + Injuries")
        g <- g + ggtitle("% Combined: Fatalities + Injuries per Event Type")
        ggplotCombinedPercent <- g + t

g <- ggplot(data=dAFatalitiesTop, aes(x=EventType, y=FATALITIES , fill=EventType))
g <- g + gb
g <- g + ylim(0, dAFatalitiesTop$FATALITIES[1]*1.2)
g <- g + coord_flip() + geom_text(aes(label=FATALITIES), hjust=-0.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab("Fatalities")
g <- g + ggtitle("Total Fatalities per Event type")
ggplotFatalities <- g + t

        g <- ggplot(data=dAFatalitiesTop, aes(x=EventType, y=percentFatalities , fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dAFatalitiesTop$percentFatalities[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percentFatalities), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% Fatalities")
        g <- g + ggtitle("% Fatalities per Event type")
        ggplotFatalitiesPercent <- g + t

g <- ggplot(data=dAInjuriesTop, aes(x=EventType, y=INJURIES , fill=EventType))
g <- g + gb
g <- g + ylim(0, dAInjuriesTop$INJURIES[1]*barMargin)
g <- g + coord_flip() + geom_text(aes(label=INJURIES), hjust=-.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab("Injuries")
g <- g + ggtitle("Total Injuries per Event type")
ggplotInjuries <- g + t

        g <- ggplot(data=dAInjuriesTop, aes(x=EventType, y=percentInjuries , fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dAInjuriesTop$percentInjuries[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percentInjuries), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% Injuries")
        g <- g + ggtitle("% Injuries per Event type")
        ggplotInjuriesPercent <- g + t



require(gridExtra)
grid.arrange(ggplotCombined, ggplotCombinedPercent,
             ggplotFatalities, ggplotFatalitiesPercent, 
             ggplotInjuries, ggplotInjuriesPercent, 
             nrow = 3, ncol=2,
             main=textGrob("Figure 1. Top 10 Total number of Fatalities, Injuries and Combined Damage per Event Type."
                           ,gp=gpar(fontsize=16,font=3))
             )


```

####12.Create Table combined Harmful (Fatalities + Injuries):

```{r tableHarmful, echo=TRUE, eval=FALSE}
#Use library xtable to generate html table
library(xtable)
xtableHarmful <- xtable(select(dataAgrCombinedTop, EventType, HealthDamage, percentHealthDamage))
print(xtableHarmful, type = "html")
```


####13.Code to generate plots that shows Top 10 Event type per Property damages, Crop damages and combined of both:

```{r plotEconomicDamages, echo=TRUE, eval=TRUE, fig.height=10, fig.width=11}
library(ggplot2)

#Common values for plots
titleSize <- 10
gb <- geom_bar(stat="identity", show_guide = FALSE, colour="black")
t <- theme(plot.title = element_text(size=titleSize))
barMargin <- 1.2
unitsDollars <- 10^9
labelDollars <- "US$ Billions"
titleSize <- 10
##Plot combined Harmful 

g <- ggplot(data=dataAgregate.EconomicDamage.Top, aes(x=EventType, y=EconomicDamage/unitsDollars, fill=EventType))
g <- g + gb
g <- g + ylim(0, dataAgregate.EconomicDamage.Top$EconomicDamage[1]/unitsDollars*barMargin)
g <- g + coord_flip() + geom_text(aes(label=round(EconomicDamage/unitsDollars,1)), hjust=-.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab(labelDollars)
g <- g + ggtitle("Event type vs. Economic Damage: Property + Crops")
ggplot.EconomicDamage <- g + t

        g <- ggplot(data=dataAgregate.EconomicDamage.Top, aes(x=EventType, y=percent.EconomicDamage , fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dataAgregate.EconomicDamage.Top$percent.EconomicDamage[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percent.EconomicDamage), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% over Total Economic Damage: Property + Crops")
        g <- g + ggtitle("Event type vs. % Economic Damage: Property + Crops")
        ggplot.EconomicDamage.Percent <- g + t
        
g <- ggplot(data=dataAgregate.PropDamage.Top, aes(x=EventType, y=PropDamage/unitsDollars, fill=EventType))
g <- g + gb
g <- g + ylim(0, dataAgregate.PropDamage.Top$PropDamage[1]/unitsDollars*barMargin)
g <- g + coord_flip() + geom_text(aes(label=round(PropDamage/unitsDollars,1)), hjust=-.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab(labelDollars)
g <- g + ggtitle("Top 10 Event type vs. Property Damage")
ggplot.PropDamage <- g + t

        g <- ggplot(data=dataAgregate.PropDamage.Top, aes(x=EventType, y=percent.PropDamage, fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dataAgregate.PropDamage.Top$percent.PropDamage[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percent.PropDamage), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% over Total Property Damage")
        g <- g + ggtitle("Top 10 Event type vs. % Property Damage")
        ggplot.PropDamage.Percent <- g + t

g <- ggplot(data=dataAgregate.CropDamage.Top, aes(x=EventType, y=CropDamage/unitsDollars, fill=EventType))
g <- g + gb
g <- g + ylim(0, dataAgregate.CropDamage.Top$CropDamage[1]/unitsDollars*barMargin)
g <- g + coord_flip() + geom_text(aes(label=round(CropDamage/unitsDollars,1)), hjust=-.25, vjust=0.5, size=4)
g <- g + xlab("Event type") + ylab(labelDollars)
g <- g + ggtitle("Top 10 Event type vs. Crop Damage")
ggplot.CropDamage <- g + t

        g <- ggplot(data=dataAgregate.CropDamage.Top, aes(x=EventType, y=percent.CropDamage, fill=EventType))
        g <- g + gb
        g <- g + ylim(0, dataAgregate.CropDamage.Top$percent.CropDamage[1]*barMargin)
        g <- g + coord_flip() + geom_text(aes(label=percent.CropDamage), hjust=-.25, vjust=0.5, size=4)
        g <- g + xlab("") + ylab("% over Total Crop Damage")
        g <- g + ggtitle("Top 10 Event type vs. % Crop Damage")
        ggplot.CropDamage.Percent <- g + t

require(gridExtra)
grid.arrange(ggplot.EconomicDamage, ggplot.EconomicDamage.Percent,
             ggplot.PropDamage, ggplot.PropDamage.Percent,
             ggplot.CropDamage, ggplot.CropDamage.Percent,
             nrow = 3, ncol=2,
             main=textGrob("Figure 2. Top 10 Event Type per Property, Crops and Combined Economic Damage."
                           ,gp=gpar(fontsize=16,font=3))
             )


```

####14.Create Table Economic Damage:

```{r tableEconomicDamage, echo=TRUE, eval=FALSE}
#Use library xtable to generate html table
library(xtable)
xtable.EconomicDamage <- xtable(select(dataAgregate.EconomicDamage.Top, EventType, EconomicDamage, percent.EconomicDamage))
print(xtable.EconomicDamage, type = "html")
```

##Results

###Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

####TOP 10 most harmful effects to poblation are:

```{r printTableHarmful, ref.label='tableHarmful', echo=FALSE, results='asis'}
```

Top 10 is `r sum(dataAgrCombinedTop$percentHealthDamage)` % of total Harmful of all events. 
Top 10 caused a total of `r as.character(total.HealthDamage.Top10)` victims.

Top 3 is `r sum(dataAgrCombinedTop$percentHealthDamage[1:3])` % of total Harmful of all events. 
Top 3 caused a total of  `r as.character(total.HealthDamage.Top3)` victims.


####Detailed graph:

```{r printPlotHarmful, ref.label='plotHarmful', echo=FALSE, fig.height=10, fig.width=11}
```


##Across the United States, which types of events have the greatest economic consequences?

####TOP 10 event type with greatest economic consequences are:

```{r printTableHarmful, ref.label='tableEconomicDamage', echo=FALSE, results='asis'}
```

Top 10 accumulate `r sum(dataAgregate.EconomicDamage.Top$percent.EconomicDamage)` % of total Harmful of all events. 
Top 10 caused a total of `r as.character(total.EconomicDamage.Top10)` US $ in economic damages.

Top 3 accumulate `r sum(dataAgregate.EconomicDamage.Top$percent.EconomicDamage[1:3])` % of total Harmful of all events. 
Top 3 caused a total of  `r as.character(total.EconomicDamage.Top3)` US $ in economic damages.


####Here's a detailed graph:

```{r printPlotEconomicDamages, ref.label='plotEconomicDamages', echo=FALSE, fig.height=10, fig.width=11}
```



